High-resolution ALMA observations of V4046\,Sgr: a circumbinary disc with a thin ring

### 2021-05-27 Draft_V4046Sgr_mnras.tex
Updates from Overleaf

% Abstract of the paper
\begin{abstract}
    The nearby V4046\,Sgr spectroscopic binary hosts a gas-rich disc known for its wide cavity and dusty ring. We present high resolution ($\sim$20--50\,mas) ALMA observations of the 1.3\,mm continuum of V4046\,Sgr which, that combined with SPHERE--IRDIS polarized images and well-sampled spectral energy distribution (SED), allow us to propose a physical model using radiative transfer (RT) predictions. The new ALMA data reveal a impressively thin ring at a radius of 13.15$\pm$0.26\,au (Ring13), with a marginally resolved radial width of 2.46$\pm$0.40\,au. Ring13 is surrounded by a $\sim$10\,au-wide gap, and it is flanked by a mm-bright outer ring (Ring24) with a sharp inner edge at 24\,au. Between 25\,au and $\sim$35\,au the brightness of Ring24 is relatively flat and then breaks into a steeper tail that reaches out to $\sim$60\,au. The observation also detects a central emission that we interpret as a tight ring made of dust grains with a lower size limit of 0.5\,mm. In order to reproduce the SED, the model that we propose requires an inner ring at $\sim$5\,au (Ring5) composed mainly of small dust grains, hiding under the IRDIS coronagraph, and surrounding an inner circumbinary disc. The surprisingly thin Ring13 is nonetheless roughly 10 times wider than its expected vertical extent. The strong near-far disc asymmetry at 1.65\,$\micron$ points at a very forward-scattering phase function, and requires grain radii of no less than 0.4\,$\micron$. The previously reported scattered-light shadow of the secondary star is also reproduced by our physical model.
\end{abstract}

@@

Ring13 is surprisingly narrow and seems to be off-centred relative to the Gaia stellar position, at the origin of coordinates in Fig.~\ref{fig:images_vs_simulated}. We determined the ring's centre and orientation by minimising the dispersion of the disc radial profiles from 6\,au to 19\,au, by using the {\tt MPolarMaps} package described in Casassus et al. (2021, in prep). The resulting disc orientation is set at a position angle (PA) of 77.31$\pm$0.03\,$\degr$ (east of north), with an inclination of 32.96$\pm$0.02\,$\degr$, and the optimal ring centre is at $\Delta \alpha = -4\pm0.02$\,mas $\Delta \delta = 13\pm0.05$\,mas relative to the Gaia position of the stars. The offset of the center of the cavity and the nominal stellar positions is entirely consistent with a pointing error, since the pointing accuracy of ALMA is $\sim$5\,mas for the signal to noise ratio of our image (calculated using the ALMA Technical Handbook).

The ring width can be measured in polar coordinates by fitting 1-D Gaussians at each azimuth, thus obtaining a width and centroid. The top panel of Fig.~\ref{fig:polarring}) shows the positions of the obtained centroids and widths over the polar decomposition of the ALMA image. On average, we obtained a FWHM of 2.83$\pm$0.25\,au, and a stellocentric radius of 13.15$\pm$0.42\,au. As the {\tt uvmem} model image has an approximate {\tt uvmem} beam of $\sim0\farcs021 \,\times \,\sim0\farcs018$, we see that Ring13 is marginally resolved. After subtraction of the {\tt uvmem} beam, the ring width is $\sim$2.46$\pm$0.40\,au.

In a second optimization of the disc orientation, but this time aiming for Ring24 with a radial domain from 20\,au to 70\,au, we obtained a PA of 76.86$\pm$0.02\,$\degr$, with an inclination of 33.92$\pm$0.01\,$\degr$ and a centre at $\Delta \alpha = 2\pm0.02$\,mas $\Delta \delta = 12\pm0.02$\,mas relative to the stars. We see that both Ring13 and Ring24 share a very similar orientation and centre, given the errors. However, quite likely due to the joint effect of all these small differences, we can see that there are some hints for a somewhat different orientation. This is summarised in the second panel of Fig.~\ref{fig:polarring}, where it shows the centroids of the Gaussian fitting at each azimuth for both optimization. The curve that follows these fitted centroids for each of the optimizations goes almost entirely within the errors of the other one, leaving the window open to the discussion of if this differences are just a consequence of the pointing error or are they a real notable feature of the disc.

@@

As the radial profiles obtained from the simulated images of the model closely resemble those deduced from the observations (Fig.~\ref{fig:radprofiles}), we can assume that the model provides a possible approximation of the disc structure, including the dimensions of Ring13. The FWHM of the gas in Ring13 in the RT model corresponds to $w_\mathrm{g} = 5.18^{+0.32}_{-0.20}$\,au, but the  large dust grains have to be radially converged to explain the ALMA image. Accordingly, taking the parametric model values for the large dust in Ring13, it has a FWHM of 2.00\,au, a radius of 13.22\,au, and a scale height FWHM of the millimetre-sized dust of 0.25\,au. The total dust mass of Ring13 would be about 0.7\,M$_{\earth}$. The model predictions for the millimetre-sized dust population in Ring13 are close to the measurements, with only a 0.07\,au difference in the centroid location of the ring, and a 0.46\,au difference in the width estimation. Since the scale height FWHM of the large grains in Ring13 is 0.25\,au, and that from the ALMA observation we can resolve a width of 2.46$\pm$0.40\,au (See sec.~\ref{subsec:ALMA}), the ring is 9.7$\pm$1.5 times more extended radially than vertically. Looking at the rest of the disc, in turn, our model reproduces the observations of Ring24 with peak intensity at $\sim$30\,au, and breaks at $\sim$36\,au. It contains a total dust mass of $\sim$45\,M$_{\earth}$.%, that stands under the lower limit of 60\,M$_{\earth}$ inferred by \citet{Ru_z_Rodr_guez_2019} from a previous 860\,$\micron$ continuum ALMA image.

@@

The key conclusions of this multi-layered analysis are as follows.
\begin{enumerate}
  \item We report the detection of a two-ringed disc structure. The narrow ring in the 1.3\,mm continuum, has a radius of 13.15$\pm$0.26\,au and an estimated width of 2.46$\pm$0.40\,au. The location of this ring is coincident with the inner ring observed in the scattered-light image. From our RT modeling we can predict that the ring includes around 0.7\,M$_{\earth}$ of millimetre-sized grains. Using the parametric model scale height FWHM value for the large grains ($H=$ 0.25\,au at 13.15\,au) we find that the ring width is roughly 10 times its estimated height. The 1.3\,mm outer ring, that starts at $\sim$24\,au and has its peak intensity at $\sim$30\,au, presents a visible break in the surface brightness at $\sim$36\,au.

>>>>

% Abstract of the paper
\begin{abstract}
    The nearby V4046\,Sgr spectroscopic binary hosts a gas-rich disc known for its wide cavity and dusty ring. We present high resolution ($\sim$20--50\,mas) ALMA observations of the 1.3\,mm continuum of V4046\,Sgr which, that combined with SPHERE--IRDIS polarized images and well-sampled spectral energy distribution (SED), allow us to propose a physical model using radiative transfer (RT) predictions. The new ALMA data reveal a impressively thin ring at a radius of 13.15$\pm$0.42\,au (Ring13), with a marginally resolved radial width of 2.46$\pm$0.20\,au. Ring13 is surrounded by a $\sim$10\,au-wide gap, and it is flanked by a mm-bright outer ring (Ring24) with a sharp inner edge at 24\,au. Between 25\,au and $\sim$35\,au the brightness of Ring24 is relatively flat and then breaks into a steeper tail that reaches out to $\sim$60\,au. The observation also detects a central emission that we interpret as a tight ring made of dust grains with a lower size limit of 0.5\,mm. In order to reproduce the SED, the model that we propose requires an inner ring at $\sim$5\,au (Ring5) composed mainly of small dust grains, hiding under the IRDIS coronagraph, and surrounding an inner circumbinary disc. The surprisingly thin Ring13 is nonetheless roughly 10 times wider than its expected vertical extent. The strong near-far disc asymmetry at 1.65\,$\micron$ points at a very forward-scattering phase function, and requires grain radii of no less than 0.4\,$\micron$. The previously reported scattered-light shadow of the secondary star is also reproduced by our physical model.
\end{abstract}

@@

Ring13 is surprisingly narrow and seems to be off-centred relative to the Gaia stellar position, at the origin of coordinates in Fig.~\ref{fig:images_vs_simulated}. We determined the ring's centre and orientation by minimising the dispersion of the disc radial profiles from 6 to 18\,au, by using the {\tt MPolarMaps} package described in Casassus et al. (2021, in prep). The resulting disc orientation is set at a position angle (PA) of 77.31$\pm$0.03\,$\degr$ (east of north), with an inclination of 32.96$\pm$0.02\,$\degr$, and the optimal ring centre is at $\Delta \alpha = -4\pm0.02$\,mas $\Delta \delta = 13\pm0.05$\,mas relative to the Gaia position of the stars. The offset of the center of the cavity and the nominal stellar positions is entirely consistent with a pointing error, since the pointing accuracy of ALMA is $\sim$5\,mas for the signal to noise ratio of our image (calculated using the ALMA Technical Handbook).

The ring width can be measured in polar coordinates by fitting 1-D Gaussians at each azimuth, thus obtaining a width and centroid. The top panel of Fig.~\ref{fig:polarring} shows the positions of the obtained centroids and widths over the polar decomposition of the ALMA image. On average, we obtained a FWHM of 2.83$\pm$0.16\,au, and a stellocentric radius of 13.15$\pm$0.42\,au. As the {\tt uvmem} model image has an approximate {\tt uvmem} beam of $\sim0\farcs021 \,\times \,\sim0\farcs018$, we see that Ring13 is marginally resolved. After subtraction of the {\tt uvmem} beam, the ring width is $\sim$2.46$\pm$0.20\,au.

In a second optimization of the disc orientation, but this time aiming for Ring24 with a radial domain from 20\,au to 70\,au, we obtained a PA of 76.86$\pm$0.02\,$\degr$, with an inclination of 33.92$\pm$0.01\,$\degr$ and a centre at $\Delta \alpha = 2\pm0.02$\,mas $\Delta \delta = 12\pm0.02$\,mas relative to the stars. We see that both Ring13 and Ring24 share a very similar orientation and centre, given the errors and the pointing error. However, quite likely due to the joint effect of all these small differences, we can see that there are some hints for a somewhat different orientation. This is summarised in the second panel of Fig.~\ref{fig:polarring}, where it shows the centroids of the Gaussian fitting at each azimuth for both optimization. The curve that follows these fitted centroids for each of the optimizations goes almost entirely within the errors of the other one, leaving the window open to the discussion of if this differences are just a consequence of the pointing error or are they a real notable feature of the disc.

@@

As the radial profiles obtained from the simulated images of the model closely resemble those deduced from the observations (Fig.~\ref{fig:radprofiles}), we can assume that the model provides a possible approximation of the disc structure, including the dimensions of Ring13. The FWHM of the gas in Ring13 in the RT model corresponds to $w_\mathrm{g} = 5.18^{+0.32}_{-0.20}$\,au, but the  large dust grains have to be radially converged to explain the ALMA image. Accordingly, taking the parametric model values for the large dust in Ring13, it has a FWHM of 2.00\,au, a radius of 13.22\,au, and a scale height FWHM of the millimetre-sized dust of 0.25\,au. The total dust mass of Ring13 would be about 0.7\,M$_{\earth}$. The model predictions for the millimetre-sized dust population in Ring13 are close to the measurements, with only a 0.07\,au difference in the centroid location of the ring, and a 0.46\,au difference in the width estimation. Since the scale height FWHM of the large grains in Ring13 is 0.25\,au, and that from the ALMA observation we can resolve a width of 2.46$\pm$0.20\,au (See sec.~\ref{subsec:ALMA}), the ring is 9.7$\pm$1.5 times more extended radially than vertically. Looking at the rest of the disc, in turn, our model reproduces the observations of Ring24 with peak intensity at $\sim$30\,au, and breaks at $\sim$36\,au. It contains a total dust mass of $\sim$45\,M$_{\earth}$.%, that stands under the lower limit of 60\,M$_{\earth}$ inferred by \citet{Ru_z_Rodr_guez_2019} from a previous 860\,$\micron$ continuum ALMA image.

@@

The key conclusions of this multi-layered analysis are as follows.
\begin{enumerate}
  \item We report the detection of a two-ringed disc structure. The narrow ring in the 1.3\,mm continuum, has a radius of 13.15$\pm$0.42\,au and an estimated width of 2.46$\pm$0.20\,au. The location of this ring is coincident with the inner ring observed in the scattered-light image. From our RT modeling we can predict that the ring includes around 0.7\,M$_{\earth}$ of millimetre-sized grains. Using the parametric model scale height FWHM value for the large grains ($H=$ 0.25\,au at 13.15\,au) we find that the ring width is roughly 10 times its estimated height. The 1.3\,mm outer ring, that starts at $\sim$24\,au and has its peak intensity at $\sim$30\,au, presents a visible break in the surface brightness at $\sim$36\,au.
